---
title: "sorl1 W1818* 6m analysis RUV"
author: "Karissa Barthelson"
date: "15/05/2020"
output: 
  html_document:
    toc: true
    toc_float: true
editor_options: 
  chunk_output_type: console
---

```{r setup, echo = T, message=FALSE}
# Set workspace -------------------------------------------------
library(magrittr)
library(reshape2)
library(tidyverse)
library(readxl)

# Analysis with bioconductor
library(limma)
library(edgeR)
library(biomaRt)
library(qvalue)
library(AnnotationHub)
library(GenomicRanges)
library(rtracklayer)
library(fgsea)
library(RUVSeq)
library(ngsReports)
library(metap)
library(harmonicmeanp)
library(cqn)
library(msigdbr)

# Visualisation
library(RColorBrewer)
library(grid)
library(gridExtra)
library(scales)
library(knitr)
library(kableExtra)
library(pheatmap)
library(ggrepel)
library(ggeasy)
library(pathview)
library(VennDiagram)
library(UpSetR)
library(pander)
library(ggfortify)
library(ggpubr)

knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)

# Viewports for grid
vp_left <- viewport(x = 0, y = 0, width = 0.5, height = 1, just = c(0, 0))
vp_right <- viewport(x = 0.5, y = 0, width = 0.5, height = 1, just = c(0,0))
```

# Introduction

Here, I am determining whether heterozygosity for the W1818* mutation in *sorl1* causes a change in the brain transcriptome. I prepared a family of fish arising from a *sorl1<sup>W1818\*/+*</sup>* x *sorl1<sup>+/+*</sup> pair mating. When this family was 6 months of age, I sacrificed 20 fish (equal males and females) and genotyped them using PCR. I then chose 6 males and 6 females, each containing 3 of each genotype for RNA extraction, library preparation and RNA-sequencing (library prep and sequencing performed by *Novogene*). 

Prior to this analysis, the provided reads (already trimmed of adaptors) were pseudo aligned to the zebrafish transcriptome using kallisto (v??) Settings were assigned as --rf-stranded  and 50 bootstraps. The index file for kallisto used the .gtf file of the GRCz11 (release 96) build of the zebrafish genome from the ENSEMBL database, with sequences for unspliced transcripts additionally included. 

## Metadata of fish

```{r meta}
metadata <- read_excel("ext_data/metadata.xlsx") %>%
  dplyr::select("Fish", "Sex", "Genotype", "SampleName", "Group", "Lane_in_sequencing", "short_sample_name") %>%
  set_colnames(c("Fish", "Sex", "Genotype", "SampleName", "Full_sample_Info", "Lane_in_sequencing", "short_sample_name")) %>%
  dplyr::filter(!is.na(SampleName)) %>%
  mutate(Sex = factor(Sex, levels = c("F", "M")),
         Genotype = factor(Genotype, levels = c("WT", "mut")),
         short_sample_name = paste0("samp_", short_sample_name))

metadata %>% 
  dplyr::select(-c(SampleName, Full_sample_Info, short_sample_name)) %>% 
  kable(caption = "Summary of fish metadata") %>% 
  kable_styling(latex_options = 'basic', full_width = F)
```
## annotations

```{r}
gr <- import.gff("ftp://ftp.ensembl.org/pub/release-96/gtf/danio_rerio/Danio_rerio.GRCz11.96.gtf.gz")

# Make a Transcript to gene mapping file
tr2gn <- gr %>% 
  subset(!is.na(transcript_id)) %>% #remove data points without a transcript id
  subset(type == "transcript") %>%
  mcols() %>%
  as.data.frame() %>%
  dplyr::select(gene_id, transcript_id) %>%
  as_tibble()

# also need GC content and length

ah <- AnnotationHub() %>%
  subset(species == "Danio rerio") %>%
  subset(rdataclass == "EnsDb")

ensDb <- ah[["AH74989"]]

grTrans <- transcripts(ensDb)

trLengths <- exonsBy(ensDb, "tx") %>%
  width() %>%
  vapply(sum, integer(1))

mcols(grTrans)$length <- trLengths[names(grTrans)]

gcGene <- grTrans %>%
  mcols() %>%
  as.data.frame() %>%
  dplyr::select(gene_id, tx_id, gc_content, length) %>%
  as_tibble() %>%
  group_by(gene_id) %>%
  summarise(
    gc_content = sum(gc_content*length) / sum(length),
    length = ceiling(median(length))
  )
grGenes <- genes(ensDb)

mcols(grGenes) %<>%
  as.data.frame() %>%
  left_join(gcGene) %>%
  as.data.frame() %>%
  DataFrame()
```

## raw counts (catchkallisto)
```{r counts}
counts <- list.files(
  path = "Kallisto_quant/",
  full.names = TRUE) %>%
  catchKallisto()

# tidy up colnames
colnames(counts$counts) <- colnames(counts$counts) %>%
  basename() %>%
  str_replace_all("(.+)Unmapped.out.mate1.fq", "samp_\\1")

# write out counts for GEO
counts$counts %>% 
  write.csv("transcriptCounts.csv")
```

## make DGE
```{r}
#Now can make the gene DGE object
geneDGE <- counts$counts %>%
  as.data.frame() %>%
  rownames_to_column("transcript_id") %>%
  dplyr::filter(!grepl("unspliced", transcript_id)) %>%
  as_tibble() %>%
  mutate(transcript_id = str_remove_all(transcript_id, "\\.[0-9]+")) %>%
  gather(key = "Sample", value = "Counts", starts_with("samp")) %>%
  left_join(tr2gn) %>%
  group_by(Sample, gene_id) %>%
  summarise(Counts = sum(Counts)) %>%
  spread(key = "Sample", value = "Counts") %>%
  column_to_rownames("gene_id") %>%
  DGEList(genes = rownames(.)) %>%
  calcNormFactors()

#Add metadata to samples
geneDGE$samples %<>% 
  as.data.frame() %>% 
  rownames_to_column("short_sample_name") %>% 
  left_join(metadata)

# Add gene info to $genes
geneDGE$genes %<>% 
  set_colnames("gene_id") %>% 
  left_join(grGenes %>% 
              as_tibble() %>% 
              dplyr::select(gene_id, gene_name, gc_content, length,
                            c(chromosome = seqnames), strand, description)) 
```
# Filtering of lowly expressed genes

Genes which are lowly expressed in RNA-seq data are not informative for differential gene expression analysis. Since one count in one RNA-seq library and 3 counts in another library does not necesserily mean a 3 fold change. Therefore we need to filter out any genes which are lowly expressed. The general rule for filtering is having a cpm above 10/(min_lib_size/1000000) in at least half of the RNA-seq libraries. 

```{r filtering_of_lowly_expressed_genes}
# filtering of lowly expressed transcripts --------------------------------

#First, i need to find the lowest library size
geneDGE$samples$lib.size %>%
  sort() %>%
  comma
min_lib_size = 13065639

#The formula for deteriming the threshold for filtering:
10/(min_lib_size/1000000) # = 0.75

keepTheseTranscripts <- (rowSums(cpm(geneDGE) > 0.75) >= 6) 

# Density plots to determine the effects of filtering

#before filtering
b4FilteringPlot <- geneDGE %>% 
  cpm(log = TRUE) %>% 
  melt %>% 
  dplyr::filter(is.finite(value)) %>% 
  ggplot(aes(x = value, colour = Var2)) +
  geom_density() + 
  guides(colour = FALSE) +
  ggtitle("Before filtering") +
  labs(x = "logCPM", y = "Proportion of Genes") 

#after filtering
geneDGE_filtered <- geneDGE[keepTheseTranscripts,, keep.lib.sizes=FALSE]

afterfilt <-geneDGE_filtered %>% 
  cpm(log = TRUE) %>% 
  melt %>% 
  dplyr::filter(is.finite(value)) %>% 
  ggplot(aes(x = value, colour = Var2)) +
  geom_density() + 
  guides(colour = FALSE) +
  ggtitle("After filtering") +
  labs(x = "logCPM", y = "Proportion of Genes") 

grid.newpage()
print(b4FilteringPlot, vp = vp_left)
print(afterfilt, vp = vp_right)
```

Peforming this filtering will leave `r dim(geneDGE_filtered$counts)[1]` out of the `r dim(geneDGE$counts)[1]` in the original dataset.

## library sizes

Library sizes after filterling lowly expressed genes. 

```{r}
geneDGE_filtered$samples %>% 
  mutate(`Library size (millions)` = lib.size/1e6) %>% 
  ggplot(aes(Full_sample_Info , `Library size (millions)`, fill = Genotype)) +
  geom_col() +
  theme_bw() +
  theme(legend.position = "none") +
  facet_wrap(~Genotype, scales = "free_x") +
  xlab("sample") +
  easy_rotate_labels(which = "x", angle = 315) 
```

# Prinicpal component analysis

We can summarise the overall similarity betweeen samples by principle component analysis (PCA). In PCA, the data (normalised counts) is subject to dimensionality-reduction by transforming the large matrix of genes and samples into a smaller set which still contains most of the information in the large set. These smaller sets are called 'principle components,' and the first prinicple component will explain the majority of the variance observed in the data. If the W1818* mutation in *sorl1* has a large effect on the brain transcriptome, we expect that the samples will seperate along the first principle component when we plot PC1 against PC2 of each sample. However, this is not what we observed as shown below. Two distinct clusters are observed, but it is not explained by genotype nor sex.

```{r}
pca_raw <- prcomp(t(cpm(geneDGE_filtered$counts, log = TRUE)))

pca_raw %>% 
   autoplot(
    data = tibble(short_sample_name = rownames(.$x)) %>%
      left_join(geneDGE_filtered$samples),
    colour = "Genotype", 
    shape = "Sex", 
    size = 4
  ) +
  theme_bw()
```

# Differential expression test - initial 

Here, I will perform a differential expression testing usig exact tests from the package *edgeR*. From this test, I will determine whether any bias for length or GC content is observed in this dataset.

```{r edegr_initial}
#need to make the genotype column in the DGE object called group
geneDGE_filtered$samples <- mutate(geneDGE_filtered$samples, 
       group = Genotype)

#Make the model matrix on the new genotype group column
design_ExactTest <- model.matrix(~group, data = geneDGE_filtered$samples)

exact_res <- geneDGE_filtered %>% 
  estimateDisp(design_ExactTest) %>% 
  edgeR::exactTest() %>% 
  topTags(n = Inf) %>% 
  .$table %>% 
  mutate(DE = FDR < 0.05) %>% 
  as_tibble()

exact_res %>% 
  head(10) %>% 
  dplyr::select(gene_id, gene_name, logFC, logCPM, PValue, FDR, everything()) %>% 
  kable(caption = 'Top 10 most DE genes via edgeR exact test') %>% 
  kable_styling(latex_options = 'basic')
```


```{r edgeR vis}
#Make a volcano plot
exact_res %>%  
  ggplot(aes(logFC, -log10(PValue), colour = DE)) +
  geom_point(alpha = 0.5, size = 2) +
  theme(aspect.ratio = 1) +
  ggtitle('Volcano plot showing any DE genes in 6 month \nold sorl1 W1818* mutant brains') +
  theme_bw() +
  scale_color_manual(values = c("grey50", "red")) +
  geom_label_repel(aes(label = gene_name), 
                   data = . %>% 
                     dplyr::filter(DE == T), 
                   show.legend = FALSE)

## MD plot
exact_res %>% 
  ggplot(aes(x = logCPM, y = logFC, colour = DE)) + 
  geom_point(alpha = 0.5, size = 2) +
  theme(aspect.ratio = 1) +
  ggtitle('MD plot') +
  theme_bw() +
  scale_color_manual(values = c("grey50", "red")) +
  geom_label_repel(aes(label = gene_name), 
                   data = . %>% 
                     dplyr::filter(DE == T), 
                   show.legend = FALSE)
```

# RUVSeq to remove unwanted variation

To assist with identifying changes to gene expression due to sorl1 genotype, I will perform RUVseq transformation to remove 1 factor of unwanted variation. 

```{r}
#Define the negative control of the 10,000 least DE genes
RUVneg <- exact_res %>% 
  as.data.frame %>%
  dplyr::arrange(desc(PValue)) %>% #sort by PValue
  .[1:10000,] %>% #Take the top 10,000 with highest P value (i.e. the least DE)
  .$gene_id

#Apply the RUV method
RUV <- geneDGE_filtered$counts %>% 
  round()  %>% 
  RUVg(RUVneg, 1) #Estimate 1 factor of unwanted variation
```

# PCA on RUV adjusted counts

After RUV-seq, some seperation across PC2 by Genotype is observed. 

```{r}
RUV$normalizedCounts %>% 
  cpm(log=TRUE) %>% 
  t() %>% 
  prcomp() %>% 
  autoplot(
    data = tibble(short_sample_name = rownames(.$x)) %>%
      left_join(geneDGE_filtered$samples),
    colour = "Genotype", 
    shape = "Sex", 
    size = 4
  ) +
  theme_bw()
```

# Differential expression with RUV covariate. 

```{r}
#Add the W_1 covariate to geneDGEfilt$samples
geneDGE_filtered$samples %<>% 
  cbind(RUV$W)

# edgeR GLM to determine the effect of RUV --------------------------------

#Design matrix including the W_1 coefficient as a covariate
designRUV <- model.matrix(~Genotype + W_1,
                          data = geneDGE_filtered$samples)

#perform the DE test includiing batch effects
glmRes_batcheffectsremoved <- estimateGLMCommonDisp(geneDGE_filtered, designRUV) %>%
  estimateGLMTagwiseDisp(designRUV) %>%
  glmFit(designRUV) %>%
  glmLRT(coef=2) %>%
  topTags(n = Inf) %>% 
  .$table %>% 
  mutate(DE = FDR < 0.05) %>% 
  as_tibble()

head(glmRes_batcheffectsremoved) %>% 
  kable(caption = 'Top 10 DE genes with batch effects removed') %>% 
  kable_styling(latex_options = 'basic')

# write out results for paper
# glmRes_batcheffectsremoved %>% 
#   dplyr::select(gene_id, gene_name, logFC, logCPM, LR, PValue, FDR, everything()) %>% 
#   write_csv("DGE_results_RUV.csv")
```

## visualisation

```{r}
glmRes_batcheffectsremoved %>% 
  ggplot(aes(x = logFC, y = -log10(PValue), colour = DE)) +
  geom_point(alpha = 0.5) +
  theme_bw() +
  theme(aspect.ratio = 1) +
  geom_label_repel(aes(label = gene_name), 
                   data = . %>% dplyr::filter(DE == T), 
                   show.legend = F) +
  scale_color_manual(values = c("grey50", "red"))
```

```{r}
glmRes_batcheffectsremoved %>% 
  ggplot(aes(x = logCPM, y = logFC, colour = DE)) + 
  geom_point(alpha = 0.5, size = 2) +
  theme(aspect.ratio = 1) +
  ggtitle('MD plot') +
  theme_bw() +
  scale_color_manual(values = c("grey50", "red")) +
  geom_label_repel(aes(label = gene_name), 
                   data = . %>% 
                     dplyr::filter(DE == T), 
                   show.legend = FALSE)

glmRes_batcheffectsremoved %>% 
  mutate(rankstat = sign(logFC)*-log10(PValue)) %>% 
  ggplot(aes(x = length, y = rankstat)) +
  geom_point(aes(colour = DE)) +
  ggtitle('rankstat v gene length') +
  theme_bw() +
  scale_x_log10() +
  scale_color_manual(values = c("grey50", "red")) +
  geom_label_repel(aes(label = gene_name), 
                   data = . %>% 
                     dplyr::filter(DE == T), 
                   show.legend = FALSE) +
  geom_hline(yintercept = 0) +
  geom_smooth(se = F)

glmRes_batcheffectsremoved %>% 
  mutate(rankstat = sign(logFC)*-log10(PValue)) %>% 
  ggplot(aes(x = gc_content, y = rankstat)) +
  geom_point(aes(colour = DE)) +
  ggtitle('rankstat v gene GC content') +
  theme_bw() +
  scale_color_manual(values = c("grey50", "red")) +
  geom_label_repel(aes(label = gene_name), 
                   data = . %>% 
                     dplyr::filter(DE == T), 
                   show.legend = FALSE) +
  geom_hline(yintercept = 0) +
  geom_smooth(se = F)
```

a small 

# Enrichment testing

Since only 2 DE genes were observed, I will perform GSEA on the entire dataset to determine any changes to pre-defined gene sets as a group. Here, I will look at the KEGG, HALLMARK and IRE gene sets. 

## Define the gene sets

```{r}
ens2Entrez <- grGenes[rownames(geneDGE_filtered)] %>%
	mcols() %>%
	as.data.frame() %>%
	as_tibble() %>%
	dplyr::select(
		gene_id, 
		entrez_gene = entrezid
	) %>%
	unchop(entrez_gene) %>%
	dplyr::filter(!is.na(entrez_gene)) 
```

```{r}
KEGG <- msigdbr("Danio rerio", category = "C2", subcategory = "CP:KEGG") %>%
	inner_join(ens2Entrez) %>%
	distinct(gs_name, gene_id, .keep_all = TRUE) %>% 
  split(f = .$gs_name) %>%
  lapply(extract2, "gene_id")

KEGG_sizes <- KEGG %>% 
  lapply(length) %>% 
  unlist %>% 
  as.data.frame() %>% 
  rownames_to_column("pathway")
```

```{r}
hallmark <- msigdbr("Danio rerio", category = "H") %>%
	inner_join(ens2Entrez) %>%
	distinct(gs_name, gene_id, .keep_all = TRUE) %>% 
  split(f = .$gs_name) %>%
  lapply(extract2, "gene_id")
```

The IRE genes were supplied already from Nhi. 

```{r}
ireGenes <- readRDS("~/Documents/sorl1_w1818x_6m/ext_data/zebrafishIreGenes.rds")
```

## perform the tests
## KEGG
### GSEA

```{r}
ranks <- glmRes_batcheffectsremoved %>% 
    mutate(PValue_withsign = sign(logFC) * log10(1/PValue)) %>% 
    arrange(PValue_withsign) %>% 
    dplyr::select(c("gene_id", "PValue_withsign")) %>% #only want the Pvalue with sign
    with(structure(PValue_withsign, names = gene_id)) %>% 
    rev()
```

```{r}
fgsea_res <- fgsea(stats = ranks, pathways = KEGG, nperm = 1e5) %>% 
      as_tibble() %>%
      dplyr::rename(FDR = padj) %>%
      mutate(padj = p.adjust(pval, "bonferroni")) %>%
      dplyr::select(pathway, pval, FDR, padj, everything()) %>%
      arrange(pval) %>%
      mutate(sig = padj < 0.05)
```

### FRY
```{r}
fry_res <- RUV$normalizedCounts %>% 
  fry(
      index = KEGG,
      design = designRUV, 
      contrast = "Genotypemut", 
      sort = "directional"
    ) %>% 
    rownames_to_column("pathway") %>% 
    as_tibble()
```

### camera

```{r}
camera_res <- RUV$normalizedCounts %>% 
  camera(
    index = KEGG,
    design = designRUV,
    contrast = 'Genotypemut', 
    inter.gene.cor = NA,
    sort = TRUE
  ) %>% 
  rownames_to_column("pathway") %>% 
  as_tibble()
```

## harmonic p

```{r}
harmonic_p_KEGG <- 
  fry_res %>% 
    mutate(fry_p = PValue.Mixed) %>% 
    dplyr::select(c(pathway, fry_p)) %>% 
    left_join(
      camera_res %>% 
        mutate(camera_p = PValue) %>% 
        dplyr::select(c(pathway, camera_p)), 
      by = "pathway"
    ) %>%
    left_join(fgsea_res %>% 
                mutate(fgsea_p = pval) %>% 
                dplyr::select(c(pathway, fgsea_p))
    ) %>% 
    nest(p = one_of(c("fry_p", "camera_p", "fgsea_p"))) %>% 
 mutate(
      harmonic_p = vapply(p, function(x){
        x <- unlist(x)
        x <- x[!is.na(x)]
        p.hmp(x, L = 3)
      }, numeric(1))
    ) %>% 
  unnest() %>% 
  mutate(harmonic_p_FDR = p.adjust(harmonic_p, "fdr"), 
         sig = harmonic_p_FDR < 0.05, 
         coef = "W1818*/+") %>% 
  arrange(harmonic_p_FDR)
```

## HALLMARK
### fgsea
```{r}
fgsea_res_hallmark <- fgsea(stats = ranks, pathways = hallmark, nperm = 1e5) %>% 
      as_tibble() %>%
      dplyr::rename(FDR = padj) %>%
      mutate(padj = p.adjust(pval, "bonferroni")) %>%
      dplyr::select(pathway, pval, FDR, padj, everything()) %>%
      arrange(pval) %>%
      mutate(sig = padj < 0.05)
```
### fry
```{r}
fry_hallmark <- RUV$normalizedCounts %>% 
  fry(
      index = hallmark,
      design = designRUV, 
      contrast = "Genotypemut", 
      sort = "directional"
    ) %>% 
    rownames_to_column("pathway") %>% 
    as_tibble()
```
## camera
```{r}
camera_hallmark <- RUV$normalizedCounts %>% 
  camera(
    index = hallmark,
    design = designRUV, 
    contrast = 'Genotypemut', 
    inter.gene.cor = NA,
    sort = TRUE
  ) %>% 
  rownames_to_column("pathway") %>% 
  as_tibble()
```
## harmonic p
```{r}
harmonic_p_hallmark <- fry_hallmark %>% 
    mutate(fry_p = PValue) %>% 
    dplyr::select(c(pathway, fry_p)) %>% 
    left_join(
      camera_hallmark %>% 
        mutate(camera_p = PValue) %>% 
        dplyr::select(c(pathway, camera_p)), 
      by = "pathway"
    ) %>%
    left_join(fgsea_res_hallmark %>% 
                mutate(fgsea_p = pval) %>% 
                dplyr::select(c(pathway, fgsea_p))
    ) %>% 
    nest(p = one_of(c("fry_p", "camera_p", "fgsea_p"))) %>% 
 mutate(
      harmonic_p = vapply(p, function(x){
        x <- unlist(x)
        x <- x[!is.na(x)]
        p.hmp(x, L = 3)
      }, numeric(1))
    ) %>% 
  unnest() %>% 
  mutate(harmonic_p_FDR = p.adjust(harmonic_p, "fdr"), 
         sig = harmonic_p_FDR < 0.05, 
         coef = "W1818*/+") %>% 
  arrange(harmonic_p_FDR) 
```

## IRE

```{r}
fgsea_res_IRE <- fgsea(stats = ranks, pathways = ireGenes, nperm = 1e5) %>% 
      as_tibble() %>%
      dplyr::rename(FDR = padj) %>%
      mutate(padj = p.adjust(pval, "bonferroni")) %>%
      dplyr::select(pathway, pval, FDR, padj, everything()) %>%
      arrange(pval) %>%
      mutate(sig = padj < 0.05)
```

```{r}
fry_IRE <- RUV$normalizedCounts %>% 
  fry(
      index = ireGenes,
      design = designRUV, 
      contrast = "Genotypemut", 
      sort = "directional"
    ) %>% 
    rownames_to_column("pathway") %>% 
    as_tibble()
```

```{r}
camera_IRE <- RUV$normalizedCounts %>% 
  camera(
    index = ireGenes,
    design = designRUV, 
    contrast = 'Genotypemut', 
    inter.gene.cor = NA,
    sort = TRUE
  ) %>% 
  rownames_to_column("pathway") %>% 
  as_tibble()
```

```{r}
harmonic_p_IRE <- fry_IRE %>% 
    mutate(fry_p = PValue) %>% 
    dplyr::select(c(pathway, fry_p)) %>% 
    left_join(
      camera_IRE %>% 
        mutate(camera_p = PValue) %>% 
        dplyr::select(c(pathway, camera_p)), 
      by = "pathway"
    ) %>%
    left_join(fgsea_res_IRE %>% 
                mutate(fgsea_p = pval) %>% 
                dplyr::select(c(pathway, fgsea_p))
    ) %>% 
    nest(p = one_of(c("fry_p", "camera_p", "fgsea_p"))) %>% 
 mutate(
      harmonic_p = vapply(p, function(x){
        x <- unlist(x)
        x <- x[!is.na(x)]
        p.hmp(x, L = 3)
      }, numeric(1))
    ) %>% 
  unnest() %>% 
  mutate(harmonic_p_FDR = p.adjust(harmonic_p, "fdr"), 
         sig = harmonic_p_FDR < 0.05, 
         coef = "W1818*/+") %>% 
  arrange(harmonic_p_FDR) 
```

# Vis

```{r}
harmonic_p_hallmark %>% 
  bind_rows(harmonic_p_IRE) %>% 
  bind_rows(harmonic_p_KEGG) %>% 
  arrange(harmonic_p_FDR) %>% 
  dplyr::filter(harmonic_p_FDR < 0.1) %>% 
  left_join(fry_hallmark %>% 
              bind_rows(fry_res) %>% 
              dplyr::select(pathway, Direction, NGenes)) %>% 
  ggplot(aes(x = -log10(harmonic_p), y = pathway)) + 
  geom_col(aes(fill = Direction)) +
  geom_vline(xintercept = 3) +
  theme_bw() +
  scale_fill_manual(values = c(brewer.pal(n = 5, name = "RdBu")[c(5,1)]))

glmRes_batcheffectsremoved %>% 
  dplyr::filter(gene_id %in% KEGG$KEGG_RIBOSOME) %>% 
  dplyr::select(gene_name, logFC) %>% 
  column_to_rownames("gene_name") %>% 
  pheatmap(cluster_cols = F, 
           color = colorRampPalette(rev(brewer.pal(n = 7, name = "RdBu")))(100)
           )

glmRes_batcheffectsremoved %>% 
  dplyr::filter(gene_id %in% KEGG$KEGG_OXIDATIVE_PHOSPHORYLATION) %>% 
  dplyr::select(gene_name, logFC) %>% 
  column_to_rownames("gene_name") %>% 
  pheatmap(cluster_cols = F, 
           color = colorRampPalette(rev(brewer.pal(n = 7, name = "RdBu")))(100)
           )

glmRes_batcheffectsremoved %>% 
  dplyr::filter(gene_id %in% KEGG$KEGG_GLYCOLYSIS_GLUCONEOGENESIS) %>% 
  dplyr::select(gene_name, logFC) %>% 
  column_to_rownames("gene_name") %>% 
  pheatmap(cluster_cols = F, 
           color = colorRampPalette(rev(brewer.pal(n = 7, name = "RdBu")))(100)
           )

glmRes_batcheffectsremoved %>% 
  dplyr::filter(gene_id %in% KEGG$KEGG_HUNTINGTONS_DISEASE) %>% 
  dplyr::select(gene_name, logFC) %>% 
  column_to_rownames("gene_name") %>% 
  pheatmap(cluster_cols = F, 
           color = colorRampPalette(rev(brewer.pal(n = 7, name = "RdBu")))(100)
           )

glmRes_batcheffectsremoved %>% 
  dplyr::filter(gene_id %in% KEGG$KEGG_PARKINSONS_DISEASE) %>% 
  dplyr::select(gene_name, logFC) %>% 
  column_to_rownames("gene_name") %>% 
  pheatmap(cluster_cols = F, 
           color = colorRampPalette(rev(brewer.pal(n = 7, name = "RdBu")))(100)
           )

glmRes_batcheffectsremoved %>% 
  dplyr::filter(gene_id %in% KEGG$KEGG_PYRUVATE_METABOLISM) %>% 
  dplyr::select(gene_name, logFC) %>% 
  column_to_rownames("gene_name") %>% 
  pheatmap(cluster_cols = F, 
           color = colorRampPalette(rev(brewer.pal(n = 7, name = "RdBu")))(100)
           )
```

## Plots for Pub

### Figure 2
```{r}
ggarrange(
  RUV$normalizedCounts %>% 
    cpm(log=TRUE) %>% 
    t() %>% 
    prcomp() %>% 
    autoplot(
      data = tibble(short_sample_name = rownames(.$x)) %>%
        left_join(geneDGE_filtered$samples %>% 
                    mutate(Genotype_for_pub = case_when(
                      Genotype == "mut" ~ "W1818*/+", 
                      Genotype == "WT" ~ "+/+"))),
      colour = "Genotype_for_pub", 
      shape = "Sex", 
      size = 4
    ) +
    theme_bw() +
    scale_color_manual(values = c("#0380fc", "#fc9d03")) +
    labs(colour = "Genotype") +
    theme(legend.position = "bottom"),
  ggarrange(
  glmRes_batcheffectsremoved %>% 
    ggplot(aes(x = logFC, y = -log10(PValue), colour = DE)) +
    geom_point(alpha = 0.5) +
    theme_bw() +
    geom_label_repel(aes(label = gene_name), 
                     data = . %>% dplyr::filter(DE == T), 
                     show.legend = F) +
    scale_color_manual(values = c("grey50", "red")) +
    labs(colour = "Differential expression?") +
    theme(legend.position = "bottom") ,
  
  glmRes_batcheffectsremoved %>% 
    ggplot(aes(x = logCPM, y = logFC, colour = DE)) + 
    geom_point(alpha = 0.5, size = 2) +
    theme_bw() +
    scale_color_manual(values = c("grey50", "red")) +
    geom_label_repel(aes(label = gene_name), 
                     data = . %>% 
                       dplyr::filter(DE == T), 
                     show.legend = FALSE) +
    labs(colour = "Differential expression?", 
         x = "Average expression (logCPM)") +
    theme( 
      legend.position = "none") ,
  common.legend = TRUE, 
  labels = c("B", "C"), 
  legend = "bottom"
  ),
  labels = c("A", " "), 
  widths = c(1, 2),
  common.legend = FALSE,
  nrow = 1) +
  ggsave("plots/Figure2.pdf", width = 20, height = 9, units = "cm", scale = 1.2, dpi = 600)
```

# Figure 3:
## harmonic p vals 
```{r}
harmonic_p_hallmark %>% 
  bind_rows(harmonic_p_IRE) %>% 
  bind_rows(harmonic_p_KEGG) %>% 
  arrange(harmonic_p_FDR) %>% 
  dplyr::filter(harmonic_p_FDR < 0.1) %>% 
  left_join(fry_hallmark %>% 
              bind_rows(fry_res) %>% 
              dplyr::select(pathway, Direction, NGenes)) %>% 
  ggplot(aes(x = -log10(harmonic_p), y = pathway)) + 
  geom_col(aes(fill = Direction, alpha = sig)) +
  scale_alpha_manual(values = c(0.5, 1)) +
  geom_vline(xintercept = 3) +
  theme_bw() +
  scale_fill_manual(values = c(brewer.pal(n = 5, name = "RdBu")[c(5,1)])) +
  ggsave("plots/harmonicP.pdf", width = 10, height = 7, units = "cm", dpi = 600, scale = 1.5)
```

### upset
```{r}
tiff("plots/Upset.tiff", width = 19, height = 10, units = "cm", res = 600, pointsize = 0.001, )
 fgsea_res %>% 
  bind_rows(fgsea_res_hallmark) %>% 
  bind_rows(fgsea_res_IRE) %>% 
  dplyr::filter(pathway %in% c(harmonic_p_hallmark %>% 
                  bind_rows(harmonic_p_KEGG) %>% 
                    bind_rows(harmonic_p_IRE) %>% 
                    dplyr::filter(harmonic_p_FDR < 0.05) %>% 
                    .$pathway)
                  ) %>% 
  dplyr::select(pathway, leadingEdge) %>% 
  unnest %>% 
  split(f = .$pathway) %>% 
  lapply(magrittr::extract2,"leadingEdge") %>% 
  fromList() %>% 
  upset(order.by = "freq", 
        nintersects = NA, 
        nsets = 6, 
        sets.x.label = "Number of genes\nin leading edge",
        set_size.numbers_size = 6
  )
dev.off()
```

# leading edge shared genes pheatmap/
```{r}
tiff("plots/LeadingEdgePheatmap.tiff", width = 24, height = 10, units = "cm", res = 0.1)
fgsea_res %>% 
  bind_rows(fgsea_res_hallmark) %>% 
  dplyr::filter(pathway %in% c("KEGG_PARKINSONS_DISEASE", "KEGG_OXIDATIVE_PHOSPHORYLATION", "HALLMARK_OXIDATIVE_PHOSPHORYLATION", "KEGG_HUNTINGTONS_DISEASE")) %>% 
  dplyr::select(pathway, leadingEdge) %>% 
  unnest %>% 
  as_tibble() %>% 
  set_colnames(c("pathway", "gene_id")) %>% 
  left_join(glmRes_batcheffectsremoved) %>% 
  dplyr::select(gene_name, pathway, logFC) %>% 
  spread(key = "pathway", value = "logFC") %>% 
  column_to_rownames("gene_name") %>% 
  t() %>% 
  pheatmap( 
    na_col = "white", 
    cluster_cols= F, 
    fontsize = 10,
   # cellwidth = 10,
    cluster_rows = F, 
    color = colorRampPalette(colors = c("darkblue", "lightblue"))(100))
dev.off()
```

## heatmaps
```{r}
tiff("plots/oxphosPheatmap.tiff", width =500, height = 1300, units = "px", res = 600)
glmRes_batcheffectsremoved %>% 
  dplyr::filter(gene_id %in% KEGG$KEGG_OXIDATIVE_PHOSPHORYLATION) %>% 
  dplyr::select(gene_name, logFC) %>% 
  column_to_rownames("gene_name") %>% 
  pheatmap(cluster_cols = F,
           treeheight_row = 0, treeheight_col = 0,
           border_color = NA,
           breaks = c(seq(min(.), 
                          0, 
                          length.out=ceiling(100/2)), 
                      seq(max(.)/100, 
                          max(.), 
                          length.out=floor(100/2))),
           color = colorRampPalette(rev(brewer.pal(n = 7, "RdBu")))(100), 
           height = 10, 
           cellwidth = 15,
           show_colnames = F, show_rownames = F
           )
dev.off()
```

### KEGG Parkinsons
```{r}
tiff("plots/KEGG_PARKINSONS_DISEASE_Pheatmap.tiff",width = 600, height = 1300, units = "px", res = 600)
glmRes_batcheffectsremoved %>% 
  dplyr::filter(gene_id %in% KEGG$KEGG_PARKINSONS_DISEASE) %>% 
  dplyr::select(gene_name, logFC) %>% 
  column_to_rownames("gene_name") %>% 
  pheatmap(cluster_cols = F,
           treeheight_row = 0, treeheight_col = 0,
           border_color = NA,
           breaks = c(seq(min(.), 
                          0, 
                          length.out=ceiling(100/2)), 
                      seq(max(.)/100, 
                          max(.), 
                          length.out=floor(100/2))),
           color = colorRampPalette(rev(brewer.pal(n = 7, "RdBu")))(100), 
           height = 10, 
           annotation_row = glmRes_batcheffectsremoved %>%
             dplyr::filter(gene_id %in% KEGG$KEGG_PARKINSONS_DISEASE) %>%
             dplyr::select(gene_name, gene_id) %>%
             mutate(in_leading_edge = case_when(
               gene_id %in% c(fgsea_res %>%
                                dplyr::filter(pathway == "KEGG_PARKINSONS_DISEASE") %>% .$leadingEdge %>% .[[1]]) ~ TRUE,
               TRUE ~ FALSE)) %>%
             dplyr::select(gene_name, in_leading_edge) %>%
             mutate(in_leading_edge = as.character(in_leading_edge)) %>% 
             column_to_rownames("gene_name"), annotation_legend = FALSE, annotation_names_row = FALSE,
           cellwidth = 15,
           annotation_colors = list(in_leading_edge = c("FALSE" = NA, "TRUE" = "grey30")),
           show_colnames = F, show_rownames = F
           )
dev.off()
```

### KEGG Huntingtons
```{r}
tiff("plots/KEGG_HUNTINGTONS_DISEASE_Pheatmap.tiff", width = 500, height = 1300, units = "px", res = 600)
glmRes_batcheffectsremoved %>% 
  dplyr::filter(gene_id %in% KEGG$KEGG_HUNTINGTONS_DISEASE) %>% 
  dplyr::select(gene_name, logFC) %>% 
  column_to_rownames("gene_name") %>% 
pheatmap(cluster_cols = F,
           treeheight_row = 0, treeheight_col = 0,
           border_color = NA,
           breaks = c(seq(min(.), 
                          0, 
                          length.out=ceiling(100/2)), 
                      seq(max(.)/100, 
                          max(.), 
                          length.out=floor(100/2))),
           color = colorRampPalette(rev(brewer.pal(n = 7, "RdBu")))(100), 
           height = 10, 
           cellwidth = 15,
           show_colnames = F, show_rownames = F
           )
dev.off()
```

### RIBOSOME

```{r}
tiff("plots/KEGG_RIBO_Pheatmap.tiff", width = 500, height = 1300, units = "px", res = 600)
glmRes_batcheffectsremoved %>% 
  dplyr::filter(gene_id %in% KEGG$KEGG_RIBOSOME) %>% 
  dplyr::select(gene_name, logFC) %>% 
  column_to_rownames("gene_name") %>% 
 pheatmap(cluster_cols = F,
           treeheight_row = 0, treeheight_col = 0,
           border_color = NA,
           breaks = c(seq(min(.), 
                          0, 
                          length.out=ceiling(100/2)), 
                      seq(max(.)/100, 
                          max(.), 
                          length.out=floor(100/2))),
           color = colorRampPalette(rev(brewer.pal(n = 7, "RdBu")))(100), 
           height = 10, 
           cellwidth = 15,
           show_colnames = F, show_rownames = F
           )
dev.off()
```

### GLYCOLYSIS
```{r}
tiff("plots/KEGG_GLYCO_Pheatmap.tiff", width = 500, height = 1300, units = "px", res = 600)
glmRes_batcheffectsremoved %>% 
  dplyr::filter(gene_id %in% KEGG$KEGG_GLYCOLYSIS_GLUCONEOGENESIS) %>% 
  dplyr::select(gene_name, logFC) %>% 
  column_to_rownames("gene_name") %>% 
 pheatmap(cluster_cols = F,
           treeheight_row = 0, treeheight_col = 0,
           border_color = NA,
           breaks = c(seq(min(.), 
                          0, 
                          length.out=ceiling(100/2)), 
                      seq(max(.)/100, 
                          max(.), 
                          length.out=floor(100/2))),
           color = colorRampPalette(rev(brewer.pal(n = 7, "RdBu")))(100), 
           height = 10, 
           cellwidth = 15,
           show_colnames = F, show_rownames = F
           )
dev.off()
```
## pathviews
```{r}
sigPaths <-harmonic_p_hallmark %>% 
  bind_rows(harmonic_p_KEGG) %>% 
  bind_rows(harmonic_p_IRE) %>% 
  dplyr::filter(harmonic_p_FDR < 0.05) %>% 
  .$pathway 
```

### oxphos
```{r}
pathview(
  species = "dre", 
  pathway.id = "dre00190", 
  gene.data = glmRes_batcheffectsremoved %>% 
    dplyr::select(gene_id, logFC) %>% 
    column_to_rownames("gene_id"), 
  gene.idtype = "ENSEMBL", 
  limit = list(gene = 0.3, 
               cpd = 0.3), 
  low = list(gene = "#2166AC", 
             cpd = "white"),
  high = list(gene = "#B2182B", 
              cpd = "white"),
  bins = list(gene = 20, 
              cpd = 20), 
  out.suffix = "_oxphos"
         )
```

## cell type check
```{r}
cell_type_markers <- readRDS("ext_data/cell_type_markers.rds")

temp <- cell_type_markers %>% 
  unlist() %>% 
  as.data.frame() %>% 
  rownames_to_column("cell_type") %>% 
  set_colnames(c("cell_type", "gene_id")) %>% 
  mutate(cell_type = cell_type %>% str_remove_all(pattern = "[0-9]+")) %>% 
  left_join(geneDGE_filtered$counts %>% 
              cpm(log = TRUE) %>% 
              as.data.frame() %>% 
              rownames_to_column("gene_id")) %>% 
  na.omit %>% 
  gather(key = "short_sample_name", value = "logCPM", 
         starts_with("samp")) %>% 
  left_join(geneDGE_filtered$samples) %>% 
  mutate(Genotype = case_when(
    Genotype == "mut" ~ "W1818*/+", 
    Genotype == "WT" ~ "+/+"
  )) %>% 
  ggplot(aes(x = Genotype, y = logCPM)) +
  geom_boxplot(aes(fill = Genotype)) +
  theme_bw() +
  facet_wrap(~cell_type) +
  scale_fill_manual(values = c("#00aaff", "#ff9d00"))  +
  theme(legend.position = "bottom") +
  ggsave("plots/celltypecheck.png", width = 6, height = 7, units = "cm", dpi = 600, scale = 1.5)

saveRDS(temp, "~/Box Sync/cellType.rds")

```

# Promotor analysis on the most DE genes using `homer`

`homer` was run using the [findMotifs.pl](http://homer.ucsd.edu/homer/motif/index.html) method to obtain the enriched known zebrafish promoter motifs and functional enrichment. Note that the background set of genes used when running `homer` is the `r dim(geneDGE_filtered)[1]` i.e. the number of genes detected in this RNA-seq experiment. 

The input genes at this stage will be the top 100 most DE genes in the RNAseq experiment. If I do not find anything, then I will increase this number a little. 

```{r}
# output the background genes:
rownames(geneDGE_filtered) %>% 
  as.data.frame() %>% 
  write_tsv("~/homer/backgroundGenes.txt")

# Get the top 100 most DE genes
glmRes_batcheffectsremoved[1:100,1] %>% 
  as.data.frame() %>% 
  write_tsv("~/homer/top100DEgenes.txt")

```

This is the command i used to generate the `homer` results:

`findMotifs.pl top100DEgenes.txt zebrafish res/ -start -1500 -end 200  -bg backgroundGenes.txt`

The results of the `homer` analysis was saved in the folder `/res`. The table below summarises the results. No significant motifs were found to be over-represented in the top 61 most DE genes. Note that I input 100 genes, but only 61 were recognised by `homer`. 

```{r}
read_tsv("~/homer/res/knownResults.txt") %>% 
  mutate(FDR = p.adjust(`P-value`, "fdr")) %>% 
  dplyr::select(`Motif Name`, Consensus, `P-value`, FDR, everything()) %>%
  head(5) %>% 
  kable(caption = "top 5 most significant promotor motifs enrichned in the top 100 most De genes according to homer") %>% 
  kable_styling()
```



```{r}
glmRes_batcheffectsremoved[1:200,1] %>% 
  as.data.frame() %>% 
  write_tsv("~/homer/top200DEgenes.txt")

read_tsv("~/homer/res2/knownResults.txt") %>% 
  mutate(bonf = p.adjust(`P-value`, "bonferroni")) %>% 
  dplyr::select(`Motif Name`, Consensus, `P-value`, bonf, everything()) %>%
  head(5) %>% 
  write.csv("homerRes.csv")
  kable(caption = "top 5 most significant promotor motifs enrichned in the top 100 most De genes according to homer") %>% 
  kable_styling()
  
  # expression of hnf4a
geneDGE %>% 
  cpm(log = T) %>% 
  as.data.frame() %>% 
  .["ENSDARG00000021494",] %>% 
  melt() %>% 
  set_colnames(c("short_sample_name", "logCPM")) %>% 
  left_join(geneDGE_filtered$samples) %>% 
  mutate(Genotype = dplyr::recode(Genotype, mut = "W1818*/+")) %>% 
  mutate(Genotype = dplyr::recode(Genotype, WT = "+/+")) %>% 
  ggplot(aes(x = Genotype, y = logCPM)) + 
  geom_boxplot(outlier.shape = NA, 
               aes(fill = Genotype)) +
  geom_jitter(aes(shape = Sex), 
              size = 4) +
  scale_fill_manual(values = c("#0380fc", "#fc9d03")) +
  theme_bw() +
  easy_rotate_x_labels(angle = 45, side = "right") +
  ggsave("plots/hnf_logCPM.png", width = 7, height = 7, units = "cm", dpi = 600, scale = 1.5)

 msigdbr("Danio rerio", category = "C3", subcategory = "TFT:TFT_Legacy") %>% 
   dplyr::filter(gs_name == "HNF4ALPHA_Q6") %>% 
	inner_join(ens2Entrez) %>%
	distinct(gs_name, gene_id, .keep_all = TRUE) %>% 
  dplyr::select(gene_id) %>% 
  left_join(glmRes_batcheffectsremoved) %>% 
   dplyr::select(gene_name, logFC)  %>% 
   column_to_rownames("gene_name") %>% 
   pheatmap(cluster_cols = F, 
            breaks = c(seq(min(.), # set 0 as white
                           0, 
                           length.out=ceiling(100/2)), 
                       seq(max(.)/100, 
                           max(.), 
                           length.out=floor(100/2))),
            color = colorRampPalette(rev(brewer.pal(n = 7, "RdBu")))(100), 
            treeheight_row = 0, cellwidth = 10)
 
 
msigdbr("Danio rerio", category = "C3", subcategory = "TFT:TFT_Legacy") %>% 
   dplyr::filter(gs_name == "HNF4ALPHA_Q6") %>% 
	inner_join(ens2Entrez) %>%
	distinct(gs_name, gene_id, .keep_all = TRUE) %>% 
  dplyr::select(gene_id) %>% 
  left_join(glmRes_batcheffectsremoved) %>% 
   dplyr::select(gene_name, logFC) %>% 
  mutate(temp = case_when(logFC > 0.5 ~ "up", 
    logFC < -0.5 ~ "down", 
    TRUE ~ "logFC ~ 0") %>% as.factor()) %>% 
  ggplot(aes(x = temp)) +
  geom_bar(aes(fill = temp), position = "stack") +
  theme_bw()

RUV$normalizedCounts %>% 
  .[c(msigdbr("Danio rerio", category = "C3", subcategory = "TFT:TFT_Legacy") %>% 
   dplyr::filter(gs_name == "HNF4ALPHA_Q6") %>% 
	inner_join(ens2Entrez) %>%
	distinct(gs_name, gene_id, .keep_all = TRUE) %>% .$gene_id),] %>% 
  t() %>% 
  prcomp() %>% 
  autoplot( data = tibble(short_sample_name = rownames(.$x)) %>%
      left_join(geneDGE_filtered$samples),
    colour = "Genotype", 
    shape = "Sex", 
    size = 4
  ) +
  scale_colour_manual(values = c("#0380fc", "#fc9d03")) +
  theme_bw() +
  ggsave("plots/hnf_PCA.png", width = 7, height = 7, units = "cm", dpi = 600, scale = 1.5)

```

## check for GC and length bias

A small bias for GC is observed, but i think it is nothing too bad. It seems at its worst at the extremes, where low number of genes have this. 

```{r}
glmRes_batcheffectsremoved %>% 
  mutate(rankstat = sign(logFC)*-log10(PValue)) %>% 
  ggplot(aes(x = length, y = rankstat)) +
  geom_point(aes(colour = DE), 
             alpha = 0.5) +
  scale_color_manual(values = c("grey50", "red")) +
  geom_smooth(se = F) +
  theme_bw() +
  scale_x_log10()

glmRes_batcheffectsremoved %>% 
  mutate(rankstat = sign(logFC)*-log10(PValue)) %>% 
  ggplot(aes(x = gc_content, y = rankstat)) +
  geom_point(aes(colour = DE), 
             alpha = 0.5) +
  scale_color_manual(values = c("grey50", "red")) +
  geom_hline(yintercept = 0) +
  geom_smooth(se = F) +
  theme_bw()
```


